---
title: "Longer simulation-based forecasting with relative R0"
output: 
  html_document:
      toc: true
      toc_float: true
      code_folding: show
vignette: >
  %\VignetteIndexEntry{Longer simulation-based forecasting with relative R0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Note: this vignette builds on the `gam-forecasting-workflow.Rmd` vignette, focusing on longer term simulations, projecting cases using the fitted GAM at each time step, then uses these cases to generate new values of susceptibles and lagged cases to project the next step.

Some other changes:  
* The GAM is now more constrained, removing the `s()` terms for fixed terms, which should be reasonable given the linear relationship between R0 and rR0 scaling.
* A log link has been added to ensure outcome is reflected as $log(E[I_t])$
* A log transform has been added to $R_0$.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE, message = FALSE
)
```

```{r setup, include = FALSE}
library(climateR0)
library(mgcv)
library(dplyr)
library(ggplot2)
library(MASS)
library(patchwork)
```

One potential use of the {climateR0} package is to use temperature-dependent R0 values as an input within a short-term forecasting model. In this vignette we outline an example using a generalised additive model, implemented in the `mgcv` package. 

## Fiji 2013/14 case study

As a case study, we'll use the same data as in the 'Getting Started' vignette. We will forecast confirmed dengue cases from the 2013/4 DENV-3 outbreak in Fiji. Here we show laboratory confirmed cases over time in Central Division.

```{r}
input_data <- fiji_2014

# Adjust for FP reporting
#input_data$cases <- round(input_data$cases/(input_data$reporting_sites/input_data$total_sites))

input_cases <- input_data |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x = date, y = cases), col = "#016c59") +
  ggplot2::scale_x_date(breaks = "month", date_labels = "%b-%y") +
  ggplot2::labs(x = "Date", y = "Cases") +
  ggplot2::theme_classic()

input_cases

```
## Set up data processing functions and forecasting model
First we set up the main functions required for our analysis:

1. `prepare_data()` - format incidence data. This includes calculating temperature-dependent R0 values using mean temperature, and adding recent lagged cases (as proxy for serial interval) and cumulative incidence (as proxy for immunity).
2. `weight_cases()` - function to weight lagged cases by the dengue serial interval.
3. `predict_cases()` -  uses a step-by-step generative model based on a input fitted GAM, with bootstrap parameter uncertainty
4. `generate_prediction()` - generates predictions for a specific forecasting window, by fitting a GAM to early data then using the `predict_cases` function to forecast
5. `define_prediction_starts()` - generates a range of possible forecast start dates, based on range of data and defined forecast horizon
6. `find_closest_days()` - function to find closest matching day of the year in vec2 for values in vec1, e.g. to allow data rows to be filled into future for `av_temp`.
 

```{r}
# Function to prepare data
prepare_data <- function(input_data, # Input data with date, cases and av_temp columns
                         pop, # Assumed population size
                         rep_prop # Assume proportion of infections reported as cases
                         ){
  
  # Check that reporting assumption does not result in more infections than pop size
  if(sum(input_data$cases, na.rm = T)/rep_prop > pop){
    stop("Assumed reporting too low, implies more infections than population size")
  }

  # Add relative R0 column
  input_data$rR0 <- temperature_r0(input_data$av_temp, "AeaeDENV")

  # Add columns tracking recent lagged cases (for serial interval)
  # and cumulative cases (for estimating immunity)
  model_data <- input_data |> 
    dplyr::mutate(date_numeric = as.numeric(date - min(date) + 1)) |> 
    dplyr::arrange(date) |> 
    dplyr::mutate(log_rR0 = log(rR0 + 0.0001),
           cumulative_cases = lag(cumsum(cases),1), 
           pop_susceptible = 1 - pmax(0, (cumulative_cases/pop)/rep_prop),
           weighted_lagged_cases = weight_cases(cases),
           log_weighted_lagged_cases = log(weighted_lagged_cases + 0.0001),
           log_pop_susceptible = log(pop_susceptible + 0.0001))
  
  # Set cumulative cases to zero initially
  model_data$cumulative_cases[1] <- 0
  model_data$pop_susceptible[1] <- 1
  model_data$log_pop_susceptible[1] <- log(1 + 0.0001)
  
  # Note the below is constrained to match the temperature date range available
  # It also includes a buffer of 5 weeks to account for serial interval calculation above
  prediction_dates <- seq(from = (min(input_data$date) + 5*7), 
                          to = max(input_data$date), by = 7)
  
  list(model_data = model_data, prediction_dates = prediction_dates)

}

# Function to weight past cases by dengue serial interval
# From Siraj et al 2017
# Note: have edited to only include 4 weeks of lag (i.e. no lag 5)

weight_cases <- function(x){
  weighted_cases <- lag(x,2)*0.2 + lag(x,3)*0.425 + lag(x,4)*0.375 
  return(weighted_cases)
}

# Function to iterate over horizon window
predict_cases <- function(train_data,
                          train_test_data,
                          prediction_bootstrap,
                          gam_mod,
                          pop,
                          rep_prop,
                          n_sim = 100, # Number of simulations
                          fix_susceptible
                          ){
  
  # Define prediction window
  horizon_start <- nrow(train_data[,,1]) + 1
  horizon_end <- nrow(train_test_data[,,1])
  
  # Set up progress bar
  pb <- txtProgressBar(horizon_start,horizon_end,style=3,title="Simulating: ")

  for (kk in horizon_start:horizon_end){
    
    # Iterate over simulation particles
    for (jj in 1:n_sim) {
      
      # Generate draw from bootstrap dataset
      index_sample <- sample(1:n_sim,1)
      train_test_data_kk <- data.frame(prediction_bootstrap[1:kk,,index_sample])
      
      # Predicted means based on original model
      original_means <- predict(gam_mod, newdata = train_test_data_kk, type = "link")
      
      # Get the variance-covariance matrix of the model coefficients
      vcov_matrix <- vcov(gam_mod)
      
      # Extract the coefficients of the model
      coef_mean <- coef(gam_mod)
      
      # Extract the dispersion parameter (theta) if negative binomial
      # theta <- gam_mod$family$getTheta()

      # Draw a new set of coefficients from the multivariate normal distribution
      coef_sim <- mvrnorm(1, mu = coef_mean, Sigma = vcov_matrix)
      
      # Compute the linear predictor with the new coefficients
      design_matrix <- model.matrix(gam_mod, newdata = train_test_data_kk)
      
      # Specify prediction depending on whether susceptible term has a coefficient
      if(fix_susceptible==TRUE){
        linear_predictor <- design_matrix %*% coef_sim + 
                            train_test_data_kk$log_rR0 + 
                            train_test_data_kk$log_weighted_lagged_cases + 
                            train_test_data_kk$log_pop_susceptible
      }else{
        linear_predictor <- design_matrix %*% coef_sim + 
                            train_test_data_kk$log_rR0 + 
                            train_test_data_kk$log_weighted_lagged_cases
      }
      
      # Transform to the response scale (mean counts) using the inverse log-link
      predicted_means <- exp(linear_predictor)
      
      # Generate simulated counts from the expectation
      # Note: we use expectation because interested in underlying dynamics
      new_simulation <- predicted_means #exp(original_means)

      # Record full range of simulated cases
      if(kk==horizon_start){
        prediction_bootstrap[1:kk,"cases_sim",jj] <- new_simulation
      }else{
        prediction_bootstrap[kk,"cases_sim",jj] <- tail(new_simulation,1)
      }
      
      # Record new step simulated cases only
      prediction_bootstrap[kk,"cases",jj] <- tail(new_simulation,1)

      # Update other outputs
      train_test_new <- data.frame(prediction_bootstrap[,,jj])
      
      cases_all_na <- train_test_new$cases[!is.na(train_test_new$cases)]
      cumulative_cases = sum(cases_all_na)
      pop_susceptible = 1 - pmin(1, (cumulative_cases/pop)/rep_prop)
      
      weighted_lagged_cases = tail(weight_cases(c(train_test_new$cases[1:kk],0)),1) # Add padding to calculate
      log_weighted_lagged_cases = log(weighted_lagged_cases + 0.0001)
      log_pop_susceptible = log(pop_susceptible + 0.0001)
      
      if(kk<horizon_end){
        prediction_bootstrap[kk+1,"log_weighted_lagged_cases",jj] <- tail(log_weighted_lagged_cases,1)
        prediction_bootstrap[kk+1,"log_pop_susceptible",jj] <- tail(log_pop_susceptible,1)
      }
      
    }
        
    # Display progress
    setTxtProgressBar(pb, kk)

  }
  
  close(pb) # Close bar
  
  prediction_bootstrap

}

# Function to generate prediction
generate_prediction <- function(model_data,
                                input_data,
                                prediction_start,
                                horizon,
                                pop,
                                rep_prop,
                                n_sim = 1, # Number of bootstrap samples
                                model_in = NULL,
                                fix_susceptible = FALSE
                                ){
  
  # DEBUG: model_data = model_data_2024; input_data = fp_incidence

  message(paste0("Generating forecasts for prediction week starting: "),
          as.Date(as.numeric(prediction_start), origin = "1970-01-01"))
  train_data <- model_data |> 
    dplyr::filter(date <= prediction_start)
  
  if(nrow(train_data)<10){
    warning("Fewer than 10 weeks of training data, results may be unstable")
  }
  
  # Run GAM model, or use pre-existing input model
  # NOTE: Ideally edit to brms so can input priors
  if(is.null(model_in)){
    # Check whether susceptible term is fixed (i.e. SIR-like) or flexible
    if(fix_susceptible==TRUE){
      gam_mod <- mgcv::gam(
                  formula = cases ~ offset(log_rR0 + 
                                    log_weighted_lagged_cases +
                                    log_pop_susceptible),
                  family = nb(link = "log"),
                  data = train_data)
    }else{
      gam_mod <- mgcv::gam(
                  formula = cases ~ offset(log_rR0 + 
                                    log_weighted_lagged_cases) +
                                    log_pop_susceptible,
                  family = nb(link = "log"),
                  data = train_data)
    }
  }else{
    gam_mod <- model_in
  }
  
  # DEBUG: test_pred <- data.frame(log_rR0 = log(0.5), log_weighted_lagged_cases=log(100), log_pop_susceptible=log(1)); predict(gam_mod,test_pred, type = "response")

  # Define rR0 for forecast period
  get_rR0_forecast <- model_data$rR0[model_data$date>as.Date(prediction_start)][1:horizon]
   
  # Set up testing rows
  prediction_rows <- data.frame(
      date = as.Date(prediction_start, origin = "1970-01-01") + seq(7,7*horizon, by = 7),
      date_numeric = max(train_data$date_numeric) + seq(7,7*horizon, by = 7),
      rR0 = get_rR0_forecast,
      log_rR0 = log(get_rR0_forecast + 0.0001),
      log_pop_susceptible = rep(tail(train_data$log_pop_susceptible,1), horizon),
      log_weighted_lagged_cases = rep(tail(train_data$log_weighted_lagged_cases,1), horizon),
      cases = rep(NA, horizon)
    )
   
  # Set up variables for first week of forecast
  cumulative_cases <- sum(train_data$cases)
  pop_susceptible <- 1 - pmax(0, (cumulative_cases/pop)/rep_prop)
  prediction_rows$log_pop_susceptible <- log(pop_susceptible + 0.0001)
  
  weighted_lagged_cases = tail(weight_cases(c(train_data$cases,0)),1)
  prediction_rows$log_weighted_lagged_cases <- log(weighted_lagged_cases + 0.0001)

  # Define training and test data
  train_test_data <-
    rbind(train_data |> dplyr::select(date, date_numeric, rR0, log_rR0,
                                      log_pop_susceptible, log_weighted_lagged_cases, cases),
          prediction_rows)
  
  # Add simulation column
  train_test_data$cases_sim <- NA
  
  # Set up array for storage
  prediction_bootstrap <- replicate(n_sim,train_test_data,simplify=F)
  
  prediction_bootstrap <- lapply(prediction_bootstrap,
                               function(df){
                                 data.frame(lapply(df,as.numeric))
                               })
  
  prediction_bootstrap <- abind::abind(prediction_bootstrap,along=3)
  
  prediction_out <- predict_cases(train_data, # Training data
                                  train_test_data, # Training + test window
                                  prediction_bootstrap, # Array for storage
                                  gam_mod, # Fitted GAM
                                  pop, # Population size
                                  rep_prop, # Reported fraction
                                  n_sim = n_sim, # Number of simulations
                                  fix_susceptible # SIR-like model = T/F
                            )
  
  # Return prediction and model
  list(prediction_out = prediction_out, model_out = gam_mod)

}

# Function to convert prediction dates and horizon into prediction values to iterate over
define_prediction_starts <- function(prediction_dates, # Dates in the dataset
                                     horizon, # Forecast horizon
                                     n_start # Number of start dates to choose
                                     ){

  # Define possible prediction start points (has to be <= horizon)
  if(length(prediction_dates)<horizon){
    stop("Need horizon to be smaller than max available prediction window")
  }
  
  # Select start dates evenly spread across prediction_dates
  pick_indices <- function(vec, x){
    if(x > length(vec)){
      stop("Number of chosen start dates can't be larger than possible set of options.")
    }
    indices <- round(seq(1, length(vec), length.out = (x+1) ))
    indices <- indices[-1] # Remove 1st option as less reliable for fitting
    vec[indices]
  }
  
  # Choose range of indices to define the end of the real-time fitting period,
  # and start of prediction period
  prediction_range <- prediction_dates[1:(length(prediction_dates)-horizon)]
  prediction_vals <- pick_indices(prediction_range, n_start)
  
  # Check prediction values value
  if(sum(is.na(prediction_vals))>0){
    stop("Need smaller horizon or stop predictions earlier")
  }
  
  prediction_vals

}

# Function to find closest matching day of the year in vec2 
# for values in vec1
find_closest_days <- function(vec1,vec2){

  doy1 <- as.numeric(format(vec1, "%j"))
  doy2 <- as.numeric(format(vec2, "%j"))
  
  sapply(doy1, function(doy){
   closest_index <- which.min(abs(doy2 - doy))
   closest_index
  })
}

```

## Fiji 2014 dengue forecast

Now we format the Fiji 2014 data for forecasting and set our forecast horizon (how many weeks ahead we want to forecast) and the dates we would like our forecasts to start from.

```{r}
population_central_division_fiji <- 342000
proportion_reported <- 0.1

# Prepare data for analysis
prepare_data_out <- prepare_data(input_data,
                                 pop = population_central_division_fiji, # Assumed population for Central Division
                                 rep_prop = proportion_reported # Assumed proportion of infections reported
                                 )

# Define data for the model
model_data <- prepare_data_out$model_data

# Define potential prediction date range
prediction_dates <- prepare_data_out$prediction_dates

# Specify range of potential forecast start dates
forecast_horizon <- 10
prediction_vals <- define_prediction_starts(prediction_dates, # Possible forecast end dates
                                          horizon = forecast_horizon, # Forecast horizon
                                          n_start = 9 # Number of forecast dates to explore
                                          ) 

```

With the model and data in place, we generate N-week ahead forecasts for a given prediction date. We can also look at `generate_output$model_out` to understand what the model is estimating (i.e. intercept term, which reflects R0, and dependence on susceptibility etc).
```{r, message=FALSE, warning=FALSE}
# Set seed for reproducibility
set.seed(123)  

# Store plots
plot_list <- list()

for(ii in 1:length(prediction_vals)){

  if(ii>length(prediction_dates)){
    stop("Need to choose index that is valid for prediction_dates vector")
  }
  
  prediction_start <- prediction_vals[ii]
  
  generate_output <- generate_prediction(model_data, # Model formatted data
                                  input_data = input_data, # Original data
                                  prediction_start, # Forecst start date
                                  horizon = forecast_horizon, # Forecast window
                                  pop = population_central_division_fiji,
                                  rep_prop = proportion_reported,
                                  n_sim = 50, # Number of bootstrap samples
                                  fix_susceptible = F # Constrain to be SIR-like 
                                  )
  
  # Get case predictions
  prediction_out <- generate_output$prediction_out
  cases_data <- prediction_out[,"cases_sim",]

  # Summarise estimates of cases over time
  summary_stats <- data.frame(
    date =   as.Date(prediction_out[,"date",1], origin = "1970-01-01"),
    median = apply(cases_data, 1, median),
    lower50 = apply(cases_data, 1, quantile, probs = 0.25, na.rm = T),
    upper50 = apply(cases_data, 1, quantile, probs = 0.75, na.rm = T),
    lower90 = apply(cases_data, 1, quantile, probs = 0.05, na.rm = T),
    upper90 = apply(cases_data, 1, quantile, probs = 0.95, na.rm = T)
  )
  
  # Calculate R0 from the GAM model intercept term
  r0_out <- exp(generate_output$model_out$coefficients[1])
  r0_estimate = data.frame(date = model_data$date,
                          r0_est = model_data$rR0*r0_out)
        
  # Define training window
  train_dataset <- model_data[model_data$date<=prediction_start,]
  
  p1 <- ggplot(summary_stats,
         aes(x = date)) +
        geom_ribbon(aes(ymin = lower90,
                  ymax = upper90,
                  fill="90% PI")) +
        geom_ribbon(aes(ymin = lower50,
                    ymax = upper50,
                    fill="50% PI")) +
        geom_line(aes( y= median,
                   color = "Forecast"),
          linewidth = 1
          ) +
        geom_point(data = model_data,
                   aes(x = date,
                   y = cases,
                   color = "Data (all)"),
                   ) +
        geom_point(data = train_dataset,
                 aes(x = date,
                 y = cases,
                 color = "Data (real-time)")
                 ) +
        scale_color_manual(values=c("Forecast"=rgb(0,0.5,0.8),
                                    "Data (all)" = "grey",
                                    "Data (real-time)" = "black")
                                    #"R0 estimate" = "blue")
        ) +
        scale_fill_manual(values=c("90% PI"=rgb(0.8,0.95,1),
                               "50% PI"=rgb(0.6,0.85,1))
                      ) +
        coord_cartesian(ylim=c(0,2e3)) +
        guides(fill="none",
               color="none") +
        labs(color=NULL,fill=NULL,
             title = paste0("Forecast start: ",prediction_start),
             x=NULL,
             y="cases") +
        theme_minimal() +
        theme(legend.position = c(0.8,0.8))
      
  plot_list[[ii]] <- p1

}
```
Then plot the combined plots
```{r}
combined_plots <- wrap_plots(plot_list)

print(combined_plots)

ggsave("combined_plots.pdf", plot = combined_plots,width=10,height=8)
```

### Further options for developement

* Need better handling of priors and uncertainty across multiple outbreaks, or even within outbreaks, e.g. by using {brms} to make posterior of outbreak 1 the prior for outbreak 2.

    
### References
[Cori et al, 2013](https://academic.oup.com/aje/article/178/9/1505/89262)  
[Camacho et al, 2018](https://www.thelancet.com/journals/langlo/article/PIIS2214-109X(18)30230-4/fulltext)  
[Mellor et al, 2023](https://www.nature.com/articles/s43856-023-00424-4)
[Siraj et al, 2017](https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0005797)
