---
title: "Example workflow for longer-term simulation with relative R0"
output: 
  html_document:
      toc: true
      toc_float: true
      code_folding: show
vignette: >
  %\VignetteIndexEntry{gam-forecasting-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Note: this vignette builds on the `gam-forecasting-workflow.Rmd` vignette, focusing on longer term simulations, projecting cases using the fitted GAM at each time step, then uses these cases to generate new values of susceptibles and lagged cases to project the next step.

Some other changes:  
* The GAM is now more constrained, removing the `s()` terms for fixed terms, which should be reasonable given the linear relationship between R0 and rR0 scaling
* A log link has been added to ensure outcome is reflected as $log(E[I_t])$
* Exploratory code for a `temperature_gt` that uses the Siraj et al temperature-dependent generation interval

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE, message = FALSE
)
```

```{r setup, include = FALSE}
library(climateR0)
library(mgcv)
library(dplyr)
library(ggplot2)
library(MASS)
library(patchwork)
```

One potential use of the {climateR0} package is to use temperature-dependent R0 values as an input within a short-term forecasting model. In this vignette we outline an example using a generalised additive model, implemented in the `mgcv` package. 

As a case study, we'll use the same data as in the 'Getting Started' vignette. We will forecast confirmed dengue cases from the 2013/4 DENV-3 outbreak in Fiji. Here we show laboratory confirmed cases over time in Central Division.

```{r}
input_data <- fiji_2014 # tahiti_2013

# Adjust for FP reporting
#input_data$cases <- round(input_data$cases/(input_data$reporting_sites/input_data$total_sites))

input_cases <- input_data |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x = date, y = cases), col = "#016c59") +
  ggplot2::scale_x_date(breaks = "month", date_labels = "%b-%y") +
  ggplot2::labs(x = "Date", y = "Cases") +
  ggplot2::theme_classic()

input_cases

```

## Model set-up

First, we calculate temperature-dependent R0 values using mean temperature.

```{r}
input_data$rR0 <- temperature_r0(input_data$av_temp, "AeaeDENV")
```

We next define a function to weight lagged cases by the dengue serial interval.

```{r}
# Function to weight past cases by dengue serial interval
# From Siraj et al 2017
# Note: have edited to only include 4 weeks of lag (i.e. no lag 5)

weight_cases <- function(x){
  weighted_cases <- lag(x,2)*0.2 + lag(x,3)*0.425 + lag(x,4)*0.375 
  return(weighted_cases)
}

```

Now we format the Fiji 2014 data for forecasting and set our forecast horizon (how many weeks ahead we want to forecast) and the dates we would like our forecasts to start from.

```{r}
pop <- 378284 # Fiji
#pop <- 190000 # FP
rep_prop <- 0.1 # Proportion of infections reported as cases

if(sum(input_data$cases)/rep_prop > pop){
  stop("Assumed reporting too low, implies more infections than population size")
}

model_data <- input_data |> 
  dplyr::mutate(date_numeric = as.numeric(date - min(date) + 1)) |> 
  dplyr::arrange(date) |> 
  dplyr::mutate(cumulative_cases = lag(cumsum(cases),1), 
                pop_susceptible = 1 - pmax(0, (cumulative_cases/pop)/rep_prop),
         weighted_lagged_cases = weight_cases(cases),
         log_weighted_lagged_cases = log(weighted_lagged_cases + 0.001),
         log_pop_susceptible = log(pop_susceptible + 0.001))

# Set cumulative cases to zero initially
model_data$cumulative_cases[1] <- 0
model_data$pop_susceptible[1] <- 1
model_data$log_pop_susceptible[1] <- log(1 + 0.001)

# Note the below is constrained to match the temperature date range available
# It also includes a buffer of 5 weeks to account for serial interval calculation above
prediction_dates <- seq(from = (min(input_data$date) + 5*7), 
                        to = max(input_data$date), by = 7)

```

## Forecasting 

First we define a prediction function that uses a step-by-step generative model based on a fitted GAM, with bootstrap parameter uncertainty:
```{r}
# Function to iterate over horizon window
predict_cases <- function(train_data,
                          train_test_data,
                          prediction_bootstrap,
                          gam_mod,
                          n_sim = 100 # number of simulations
                          ){
  
  # Define prediction window
  horizon_start <- nrow(train_data[,,1]) + 1
  horizon_end <- nrow(train_test_data[,,1])
  
  # Set up progress bar
  pb <- txtProgressBar(horizon_start,horizon_end,style=3,title="Simulating: ")

  for (kk in horizon_start:horizon_end){
    
    # Iterate over simulation particles
    for (jj in 1:n_sim) {
      
      # Generate draw from bootstrap dataset
      index_sample <- sample(1:n_sim,1)
      train_test_data_kk <- data.frame(prediction_bootstrap[1:kk,,index_sample])
      
      # Predicted means based on original model
      original_means <- predict(gam_mod, newdata = train_test_data_kk, type = "link")
      
      # Get the variance-covariance matrix of the model coefficients
      vcov_matrix <- vcov(gam_mod)
      
      # Extract the coefficients of the model
      coef_mean <- coef(gam_mod)
      
      # Extract the dispersion parameter (theta) if negative binomial
      # theta <- gam_mod$family$getTheta()

      # Draw a new set of coefficients from the multivariate normal distribution
      coef_sim <- mvrnorm(1, mu = coef_mean, Sigma = vcov_matrix)
      
      # Compute the linear predictor with the new coefficients
      linear_predictor <- model.matrix(gam_mod, newdata = train_test_data_kk) %*% coef_sim
      
      # Transform to the response scale (mean counts) using the inverse log-link
      predicted_means <- exp(linear_predictor)
      
      # Generate simulated counts from the expectation
      # Note: we use expecation because interested in underlying dynamics
      new_simulation <- predicted_means

      # Record full range of simulated cases
      if(kk==horizon_start){
        prediction_bootstrap[1:kk,"cases_sim",jj] <- new_simulation
      }else{
        prediction_bootstrap[kk,"cases_sim",jj] <- tail(new_simulation,1)
      }
      
      # Record new step simulated cases only
      prediction_bootstrap[kk,"cases",jj] <- tail(new_simulation,1)

      # Update other outputs
      train_test_new <- data.frame(prediction_bootstrap[,,jj])
      
      cases_all_na <- train_test_new$cases[!is.na(train_test_new$cases)]
      cumulative_cases = sum(cases_all_na)
      pop_susceptible = 1 - pmin(1, (cumulative_cases/pop)/rep_prop)
      
      weighted_lagged_cases = tail(weight_cases(c(train_test_new$cases[1:kk],0)),1) # Add padding to calculate
      log_weighted_lagged_cases = log(weighted_lagged_cases + 0.001)
      log_pop_susceptible = log(pop_susceptible + 0.001)
      
      if(kk<horizon_end){
        prediction_bootstrap[kk+1,"log_weighted_lagged_cases",jj] <- tail(log_weighted_lagged_cases,1)
        prediction_bootstrap[kk+1,"log_pop_susceptible",jj] <- tail(log_pop_susceptible,1)
      }
      
    }
        
    # Display progress
    setTxtProgressBar(pb, kk)

  }
  
  close(pb) # Close bar
  
  prediction_bootstrap

}
```

Next, we step up a model function to generate predictions for a specific forecasting window, by fitting a GAM to early data then using the `predict_cases` function to forecast:
```{r}
generate_prediction <- function(model_data,
                                input_data, #fiji_2014,
                                prediction_start,
                                horizon,
                                n_sim = 1 # Number of bootstrap samples
                                ){

  message(paste0("Generating forecasts for prediction week starting: "),
          as.Date(as.numeric(prediction_start), origin = "1970-01-01"))
  train_data <- model_data |> 
    dplyr::filter(date <= prediction_start)
  
  if(nrow(train_data)<10){
    warning("Fewer than 10 weeks of training data, results may be unstable")
  }
  
  # Run GAM model
  gam_mod <- mgcv::gam(
    formula = cases ~ rR0 + (log_weighted_lagged_cases) + (log_pop_susceptible),
    family = nb(link = "log"),
    data = train_data)
   
  # Define rR0 for forecast period
  get_rR0_forecast <- input_data$rR0[input_data$date>as.Date(prediction_start)][1:horizon]
   
  # Set up testing rows
  prediction_rows <- data.frame(
      date = as.Date(prediction_start, origin = "1970-01-01") + seq(7,7*horizon, by = 7),
      date_numeric = max(train_data$date_numeric) + seq(7,7*horizon, by = 7),
      rR0 = get_rR0_forecast,
      log_pop_susceptible = rep(tail(train_data$log_pop_susceptible,1), horizon),
      log_weighted_lagged_cases = rep(tail(train_data$log_weighted_lagged_cases,1), horizon),
      cases = rep(NA, horizon)
    )
   
  # Set up variables for first week of forecast
  cumulative_cases <- sum(train_data$cases)
  pop_susceptible <- 1 - pmax(0, (cumulative_cases/pop)/rep_prop)
  prediction_rows$log_pop_susceptible <- log(pop_susceptible + 0.001)
  
  weighted_lagged_cases = tail(weight_cases(c(train_data$cases,0)),1)
  prediction_rows$log_weighted_lagged_cases <- log(weighted_lagged_cases + 0.001)

  # Define training and test data
  train_test_data <-
    rbind(train_data |> dplyr::select(date, date_numeric, rR0, 
                                      log_pop_susceptible, log_weighted_lagged_cases, cases),
          prediction_rows)
  
  # Add simulation column
  train_test_data$cases_sim <- NA
  
  # Set up array for storage
  prediction_bootstrap <- replicate(n_sim,train_test_data,simplify=F)
  
  prediction_bootstrap <- lapply(prediction_bootstrap,
                               function(df){
                                 data.frame(lapply(df,as.numeric))
                               })
  
  prediction_bootstrap <- abind::abind(prediction_bootstrap,along=3)
  
  
  prediction_out <- predict_cases(train_data,
                                  train_test_data,
                                  prediction_bootstrap,
                                  gam_mod,
                                  n_sim = n_sim # number of simulations
                            )
  
  prediction_out

}
```

With the model in place, we generate N-week ahead forecasts for a given prediction date:
```{r, echo = T}
# Set seed for reproducibility
set.seed(123)  

# Define forecast window
horizon <- 10

# Define possible prediction start points (has to be <= horizon)
if(length(prediction_dates)<horizon){
  stop("Need horizon to be smaller than max available prediction window")
}

# Choose range of indices to define the end of the real-time fitting period,
# and start of prediction period
prediction_range <- prediction_dates[1:(length(prediction_dates)-horizon)]
prediction_vals <- prediction_range[seq(4,10,2)]

# Check prediction values value
if(sum(is.na(prediction_vals))>0){
  stop("Need smaller horizon or stop predictions earlier")
}

# Store plots
plot_list <- list()

for(ii in 1:length(prediction_vals)){

  if(ii>length(prediction_dates)){
    stop("Need to choose index that is valid for prediction_dates vector")
  }
  
  prediction_start <- prediction_vals[ii]
  
  prediction_out <- generate_prediction(model_data,
                                  input_data = input_data,
                                  prediction_start,
                                  horizon,
                                  n_sim = 50 # Number of bootstrap samples
                                  )
  
  cases_data <- prediction_out[,"cases_sim",]

  summary_stats <- data.frame(
    date =   as.Date(prediction_out[,"date",1]),
    median = apply(cases_data, 1, median),
    lower50 = apply(cases_data, 1, quantile, probs = 0.25, na.rm = T),
    upper50 = apply(cases_data, 1, quantile, probs = 0.75, na.rm = T),
    lower90 = apply(cases_data, 1, quantile, probs = 0.05, na.rm = T),
    upper90 = apply(cases_data, 1, quantile, probs = 0.95, na.rm = T)
  )
  
  # Define training window
  train_dataset <- model_data[model_data$date<=prediction_start,]
  
  p1 <- ggplot(summary_stats,
         aes(x = date)) +
        geom_ribbon(aes(ymin = lower90,
                  ymax = upper90,
                  fill="90% PI")) +
        geom_ribbon(aes(ymin = lower50,
                    ymax = upper50,
                    fill="50% PI")) +
        geom_line(aes( y= median,
                   color = "Forecast"),
          linewidth = 1
          ) +
        geom_point(data = model_data,
                   aes(x = date,
                   y = cases,
                   color = "Data (all)"),
                   ) +
        geom_point(data = train_dataset,
                 aes(x = date,
                 y = cases,
                 color = "Data (real-time)")
                 ) +
        scale_color_manual(values=c("Forecast"=rgb(0,0.5,0.8),
                                    "Data (all)" = "grey",
                                    "Data (real-time)" = "black")
        ) +
        scale_fill_manual(values=c("90% PI"=rgb(0.8,0.95,1),
                               "50% PI"=rgb(0.6,0.85,1))
                      ) +
        coord_cartesian(ylim=c(0,5e3)) +
        guides(fill="none",
               color="none") +
        labs(color=NULL,fill=NULL,
             title = paste0("Forecast start: ",prediction_start),
             x=NULL,
             y="cases") +
        theme_minimal() +
        theme(legend.position = c(0.8,0.8))
      
  plot_list[[ii]] <- p1

}
```
Then plot the combined plots
```{r}
combined_plots <- wrap_plots(plot_list)


ggsave("combined_plots.pdf", plot = combined_plots,width=10,height=8)
```



### Further thoughts

* Need better handling of uncertainty in propagated outbreaks

* Potential for temperature-dependent generation interval:
```{r}
weight_cases_temp <- function(x, weight_vals){
  weighted_cases <- lag(x,2)*weight_vals[2] + 
                    lag(x,3)*weight_vals[3] + 
                    lag(x,4)*weight_vals[4] + 
                    lag(x,5)*weight_vals[5]
  return(weighted_cases)
}


temperature_gt <- function(x, temp){
  
  daily_weights <- gt.function(temp)(1:50) # From arbotemp paper
  daily_weights <- daily_weights/sum(daily_weights)
  
  weekly_weights <- tapply(daily_weights, (seq_along(daily_weights) - 1) %/% 7, sum)
  
  weight_cases_temp(x, weekly_weights)

}

```
    
### References
[Cori et al, 2013](https://academic.oup.com/aje/article/178/9/1505/89262)  
[Camacho et al, 2018](https://www.thelancet.com/journals/langlo/article/PIIS2214-109X(18)30230-4/fulltext)  
[Mellor et al, 2023](https://www.nature.com/articles/s43856-023-00424-4)
[Siraj et al, 2017](https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0005797)
