---
title: "GAM-based Simulation Forecasting with climateR0"
output: 
  html_document:
      toc: true
      toc_float: true
      code_folding: show
vignette: >
  %\VignetteIndexEntry{GAM-based Simulation Forecasting with climateR0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE, message = FALSE
)
```

```{r setup, include = FALSE}
library(climateR0)
library(ggplot2)
library(patchwork)
```

This vignette demonstrates how to use the `climateR0` package for GAM-based forecasting of disease cases, incorporating temperature-dependent R0 values. We'll use the Fiji 2013/14 dengue outbreak as a case study.

## Data Preparation

First, let's load and examine the data:

```{r}
# Load example data
data(fiji_2014)

# Plot the case data
input_cases <- ggplot2::ggplot(fiji_2014) +
  ggplot2::geom_line(ggplot2::aes(x = date, y = cases), col = "#016c59") +
  ggplot2::scale_x_date(breaks = "month", date_labels = "%b-%y") +
  ggplot2::labs(x = "Date", y = "Cases") +
  ggplot2::theme_classic()

input_cases
```

## Setting Up Parameters

We need to define some key parameters for our analysis:

```{r}
# Population and reporting parameters
pop_central_division_fiji <- 342000
proportion_reported <- 0.1

# Prepare data for analysis
prepare_data_out <- prepare_data(
  input_data = fiji_2014,
  pop = pop_central_division_fiji,
  rep_prop = proportion_reported
)

# Extract processed data
model_data <- prepare_data_out$model_data
prediction_dates <- prepare_data_out$prediction_dates
```

## Setting Up Forecast Parameters

Now we'll set up our forecast parameters:

```{r}
# Define forecast parameters
forecast_horizon <- 10  # weeks ahead to forecast
n_start_dates <- 9     # number of forecast start dates to explore

# Generate prediction start dates
prediction_vals <- define_prediction_starts(
  prediction_dates = prediction_dates,
  horizon = forecast_horizon,
  n_start = n_start_dates
)
```

## Generating Forecasts

Let's generate forecasts for each start date:

```{r}
# Set seed for reproducibility
set.seed(123)

# Store plots
plot_list <- list()

# Generate forecasts for each start date
for (ii in seq_along(prediction_vals)) {
  prediction_start <- prediction_vals[ii]

  # Generate prediction
  generate_output <- generate_prediction(
    model_data = model_data,
    input_data = fiji_2014,
    prediction_start = prediction_start,
    horizon = forecast_horizon,
    pop = pop_central_division_fiji,
    rep_prop = proportion_reported,
    n_sim = 50,
    fix_all_covariates = FALSE
  )

  # Extract predictions
  prediction_out <- generate_output$prediction_out
  cases_data <- prediction_out[, "cases_sim", ]

  # Calculate summary statistics
  summary_stats <- data.frame(
    date = as.Date(prediction_out[, "date", 1], origin = "1970-01-01"),
    median = apply(cases_data, 1, median),
    lower50 = apply(cases_data, 1, quantile, probs = 0.25, na.rm = TRUE),
    upper50 = apply(cases_data, 1, quantile, probs = 0.75, na.rm = TRUE),
    lower90 = apply(cases_data, 1, quantile, probs = 0.05, na.rm = TRUE),
    upper90 = apply(cases_data, 1, quantile, probs = 0.95, na.rm = TRUE)
  )

  # Calculate R0 from the GAM model
  r0_out <- exp(generate_output$model_out$coefficients[1])
  r0_estimate <- data.frame(
    date = model_data$date,
    r0_est = model_data$rR0 * r0_out
  )

  # Define training window
  train_dataset <- model_data[model_data$date <= prediction_start, ]

  # Create plot
  p1 <- ggplot(summary_stats, aes(x = date)) +
    geom_ribbon(aes(ymin = lower90, ymax = upper90, fill = "90% PI")) +
    geom_ribbon(aes(ymin = lower50, ymax = upper50, fill = "50% PI")) +
    geom_line(aes(y = median, color = "Forecast"), linewidth = 1) +
    geom_point(data = model_data,
               aes(x = date, y = cases, color = "Data (all)")) +
    geom_point(data = train_dataset,
               aes(x = date, y = cases, color = "Data (real-time)")) +
    scale_color_manual(values = c(
      Forecast = rgb(0, 0.5, 0.8),
      "Data (all)" = "grey",
      "Data (real-time)" = "black"
    )) +
    scale_fill_manual(values = c(
      "90% PI" = rgb(0.8, 0.95, 1),
      "50% PI" = rgb(0.6, 0.85, 1)
    )) +
    coord_cartesian(ylim = c(0, 2e3)) +
    guides(fill = "none", color = "none") +
    labs(
      title = paste0("Forecast start: ", prediction_start),
      x = NULL,
      y = "cases"
    ) +
    theme_minimal() +
    theme(legend.position = c(0.8, 0.8))

  plot_list[[ii]] <- p1
}

# Display all plots
wrap_plots(plot_list, ncol = 3)
```

## Model Interpretation

The GAM model provides insights into the relationship between cases and various factors:

```{r}
# Extract model from last prediction
last_model <- generate_output$model_out

# Print model summary
summary(last_model)
```

## Key Features of the Implementation

1. **Data Preparation**:
   - Incorporates temperature-dependent R0 values
   - Calculates weighted lagged cases using the dengue serial interval
   - Tracks population susceptibility

2. **Forecasting**:
   - Uses GAM models with negative binomial distribution
   - Incorporates uncertainty through bootstrap simulations
   - Provides prediction intervals at 50% and 90% levels

3. **Model Flexibility**:
   - Can use either SIR-like (fixed covariates) or flexible models
   - Allows for pre-fitted models to be used
   - Handles missing data appropriately

## References

- Siraj et al. (2017) for dengue serial interval estimates
- Temperature-dependent R0 calculations from the climateR0 package
