---
title: "Example workflow for longer-term simulation with relative R0"
output: 
  html_document:
      toc: true
      toc_float: true
      code_folding: show
vignette: >
  %\VignetteIndexEntry{gam-forecasting-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Note: this vignette builds on the `gam-forecasting-workflow.Rmd` vignette, focusing on longer term simulations, projecting cases using the fitted GAM at each time step, then uses these cases to generate new values of susceptibles and lagged cases to project the next step.

Some other changes:  
* The GAM is now more constrained, removing the `s()` terms for fixed terms, which should be reasonable given the linear relationship between R0 and rR0 scaling
* A log link has been added to ensure outcome is reflected as $log(E[I_t])$
* Exploratory code for a `temperature_gt` that uses the Siraj et al temperature-dependent generation interval

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE, message = FALSE
)
```

```{r setup, include = FALSE}
library(climateR0)
library(here)
library(mgcv)
library(dplyr)
library(gratia)
library(ggplot2)
library(MASS)
```

One potential use of the {climateR0} package is to use temperature-dependent R0 values as an input within a short-term forecasting model. In this vignette we outline an example using a generalised additive model, implemented in the `mgcv` package. 

As a case study, we'll use the same data as in the 'Getting Started' vignette. We will forecast confirmed dengue cases from the 2013/4 DENV-3 outbreak in Fiji. Here we show laboratory confirmed cases over time in Central Division.

```{r}
fiji_cases <- fiji_2014 |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x = date, y = cases), col = "#016c59") +
  ggplot2::scale_x_date(breaks = "month", date_labels = "%b-%y") +
  ggplot2::labs(x = "Date", y = "Cases") +
  ggplot2::theme_classic()

fiji_cases

```

## Model set-up

First, we calculate temperature-dependent R0 values using mean temperature.

```{r}
fiji_2014$rR0 <- temperature_r0(fiji_2014$av_temp, "AeaeDENV")
```

We next define a function to weight lagged cases by the dengue serial interval.

```{r}
# Function to weight past cases by dengue serial interval
# From Siraj et al 2017

weight_cases <- function(x){
  weighted_cases <- lag(x,2)*0.2 + lag(x,3)*0.425 + lag(x,4)*0.25 + lag(x,5)*0.125
  return(weighted_cases)
}

```

Now we format the Fiji 2014 data for forecasting and set our forecast horizon (how many weeks ahead we want to forecast) and the dates we would like our forecasts to start from.

```{r}
pop <- 378284
rep_prop <- 0.1 # Proportion of infections reported as cases

model_data <- fiji_2014 |> 
  dplyr::mutate(date_numeric = as.numeric(date - min(date) + 1)) |> 
  dplyr::arrange(date) |> 
  dplyr::mutate(cumulative_cases = cumsum(cases), 
                pop_susceptible = 1 - pmax(0, (cumulative_cases/pop)/rep_prop),
         weighted_lagged_cases = weight_cases(cases),
         log_weighted_lagged_cases = log(weighted_lagged_cases + 0.001),
         log_pop_susceptible = log(pop_susceptible + 0.001))

# Projection window into the future
horizon <- 10

# Note the below is constrained to match the temperature date range available
prediction_dates <- seq(from = as.Date("2013-12-30"), to = as.Date("2014-09-01"), by = 7)

```

## Forecasting 

First we define a prediction function that uses a step-by-step generative model based on a fitted GAM, with bootstrap parameter uncertainty:
```{r}
# Function to iterate over horizon window
predict_cases <- function(train_data,
                          prediction_bootstrap,
                          gam_model,
                          n_sim = 100 # number of simulations
                          ){
  
  # Define prediction window
  horizon_start <- nrow(train_data[,,1]) + 1
  horizon_end <- nrow(train_test_data[,,1])
  
  for (kk in horizon_start:horizon_end){
    
    # Iterate over simulation particles
    for (jj in 1:n_sim) {
      
      # Generate draw from bootstrap dataset
      index_sample <- sample(1:n_sim,1)
      train_test_data_kk <- data.frame(prediction_bootstrap[1:kk,,index_sample])
      
      # Predicted means based on original model
      original_means <- predict(gam_mod, newdata = train_test_data_kk, type = "link")
      
      # Get the variance-covariance matrix of the model coefficients
      vcov_matrix <- vcov(gam_mod)
      
      # Extract the coefficients of the model
      coef_mean <- coef(gam_mod)
      
      # Extract the dispersion parameter (theta) if negative binomial
      # theta <- gam_mod$family$getTheta()

      # Draw a new set of coefficients from the multivariate normal distribution
      coef_sim <- mvrnorm(1, mu = coef_mean, Sigma = vcov_matrix)
      
      # Compute the linear predictor with the new coefficients
      linear_predictor <- model.matrix(gam_mod, newdata = train_test_data_kk) %*% coef_sim
      
      # Transform to the response scale (mean counts) using the inverse log-link
      predicted_means <- exp(linear_predictor)
      
      # Generate simulated counts from the negative binomial distribution
      # Suppress NA warnings (because early counts will always be NA, so fine to return NA)
      new_simulation <- predicted_means
      # new_simulation <- suppressWarnings(rnbinom(n = length(predicted_means), 
      #                                            mu = predicted_means, 
      #                                            size = theta))
      
      if(kk==horizon_start){
        prediction_bootstrap[1:kk,"cases",jj] <- new_simulation
      }else{
        prediction_bootstrap[kk,"cases",jj] <- tail(new_simulation,1)
      }

      # Update other outputs
      train_test_new <- data.frame(prediction_bootstrap[,,jj])
      
      cases_all_na <- train_test_new$cases[!is.na(train_test_new$cases)]
      cumulative_cases = sum(cases_all_na)
      pop_susceptible = 1 - pmin(1, (cumulative_cases/pop)/rep_prop)
      weighted_lagged_cases = weight_cases(cases_all_na)
      log_weighted_lagged_cases = log(weighted_lagged_cases + 0.001)
      log_pop_susceptible = log(pop_susceptible + 0.001)
      
      if(kk<horizon_end){
        prediction_bootstrap[kk+1,"log_weighted_lagged_cases",jj] <- tail(log_weighted_lagged_cases,1)
        prediction_bootstrap[kk+1,"log_pop_susceptible",jj] <- tail(log_pop_susceptible,1)
      }
      
    }

  }
  
  prediction_bootstrap

}
```

We now generate N-week ahead forecasts for a given prediction date:
```{r, echo = T}

n_sim <- 20 # Number of bootstrap samples
set.seed(123)  # For reproducibility

all_forecasts <- NULL 
weekly_forecasts <- NULL

# Define possible prediction start points (has to be <= horizon)
prediction_range <- prediction_dates[1:(length(prediction_dates)-horizon)]


#for(ii in 1:length(prediction_range)){ 
  
# ADD DEBUG: message to check timeseries lengths (see rR0 below)
ii=8

prediction_start <- prediction_range[ii]

message(paste0("Generating forecasts for prediction week starting: "), as.Date(as.numeric(prediction_start), origin = "1970-01-01"))
train_data <- model_data |> 
  dplyr::filter(date <= prediction_start)

# Run GAM model
 gam_mod <- mgcv::gam(
  formula = cases ~ rR0 + (log_weighted_lagged_cases) + (log_pop_susceptible),
  family = nb(link = "log"),
  data = train_data)
 
 # Define rR0 for forecast period
 get_rR0_forecast <- fiji_2014$rR0[fiji_2014$date>as.Date(prediction_start)][1:horizon]
 
 # Set up testing rows
prediction_rows <- data.frame(
  date = as.Date(prediction_start, origin = "1970-01-01") + seq(7,7*horizon, by = 7),
  date_numeric = max(train_data$date_numeric) + seq(7,7*horizon, by = 7),
  rR0 = get_rR0_forecast,
  log_pop_susceptible = rep(tail(train_data$log_pop_susceptible,1), horizon),
  log_weighted_lagged_cases = rep(tail(train_data$log_weighted_lagged_cases,1), horizon),
  cases = rep(NA, horizon)
)

# Define training and test data
train_test_data <-
  rbind(train_data |> dplyr::select(date, date_numeric, rR0, log_pop_susceptible, log_weighted_lagged_cases, cases),
        prediction_rows)

# Set up array for storage
prediction_bootstrap <- replicate(n_sim,train_test_data,simplify=F)

prediction_bootstrap <- lapply(prediction_bootstrap,
                             function(df){
                               data.frame(lapply(df,as.numeric))
                             })

prediction_bootstrap <- abind::abind(prediction_bootstrap,along=3)


prediction_out <- predict_cases(train_data,
                          prediction_bootstrap,
                          gam_model,
                          n_sim = n_sim # number of simulations
                          )

cases_data <- prediction_out[,"cases",]

#col_grad <- colorRampPalette(c("orange", "blue"))(length(prediction_range))

summary_stats <- data.frame(
  date = train_test_data$date,
  median = apply(cases_data, 1, median),
  lower50 = apply(cases_data, 1, quantile, probs = 0.25, na.rm = T),
  upper50 = apply(cases_data, 1, quantile, probs = 0.75, na.rm = T),
  lower90 = apply(cases_data, 1, quantile, probs = 0.05, na.rm = T),
  upper90 = apply(cases_data, 1, quantile, probs = 0.95, na.rm = T)
)



p1 <- ggplot(summary_stats,
       aes(x = date)) +
      geom_ribbon(aes(ymin = lower90,
                  ymax = upper90),
              fill="lightblue",
              alpha = 0.4) +
  geom_ribbon(aes(ymin = lower50,
                  ymax = upper50),
              fill="lightblue",
              alpha = 0.6) +

  geom_line(aes( y= median),
            color = "darkcyan",
            linewidth = 1) +
  geom_point(data = model_data,
             aes(x = date,
             y = cases),
             color = "grey") +
  geom_point(data = train_data,
           aes(x = date,
           y = cases),
           color = "black") +
  coord_cartesian(ylim=c(0,2e3)) +
  theme_minimal()

p1
```




### Further thoughts

* Need better handling of uncertainty in propagated outbreaks

* Potential for temperature-dependent generation interval:
```{r}
weight_cases_temp <- function(x, weight_vals){
  weighted_cases <- lag(x,2)*weight_vals[2] + 
                    lag(x,3)*weight_vals[3] + 
                    lag(x,4)*weight_vals[4] + 
                    lag(x,5)*weight_vals[5]
  return(weighted_cases)
}


temperature_gt <- function(x, temp){
  
  daily_weights <- gt.function(temp)(1:50) # From arbotemp paper
  daily_weights <- daily_weights/sum(daily_weights)
  
  weekly_weights <- tapply(daily_weights, (seq_along(daily_weights) - 1) %/% 7, sum)
  
  weight_cases_temp(x, weekly_weights)

}

```
    
### References
[Cori et al, 2013](https://academic.oup.com/aje/article/178/9/1505/89262)  
[Camacho et al, 2018](https://www.thelancet.com/journals/langlo/article/PIIS2214-109X(18)30230-4/fulltext)  
[Mellor et al, 2023](https://www.nature.com/articles/s43856-023-00424-4)
[Siraj et al, 2017](https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0005797)
